---
title: "Lab 4"
author: "Allison Louie"
format:
  html:
    embed-resources: true
---
## #1 Read in the data
```{r}
library(ggplot2)
library(dplyr)
library(leaflet)
library(R.utils)

if (!file.exists("met_all.gz"))
 download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
    )
met <- data.table::fread("met_all.gz")
```

## #2 Prepare the data
```{r}
met <- met[met$temp >= -17][elev == 9999.0, elev := NA]
met[, week := as.integer(format(as.Date(paste(year, month, day, sep = "-")), "%U"))]
met <- met[week == 31 | week == 32]

met_avg <- met[,.(
temp = mean(temp,na.rm=TRUE),
rh = mean(rh,na.rm=TRUE),
wind.sp = mean(wind.sp,na.rm=TRUE),
vis.dist = mean(vis.dist,na.rm=TRUE),
dew.point = mean(dew.point,na.rm=TRUE),
lat = mean(lat),
lon = mean(lon),
elev = mean(elev,na.rm=TRUE)
), by=c("USAFID", "day")
]

met_avg = mutate(met_avg, region = ifelse(lon >= -98 & lat >= 39.71, "NE",
                            ifelse(lon >= -98 & lat < 39.71, "NW",
                            ifelse(lon < -98 & lat >= 39.71, "SE", "SW"))))

met_avg = mutate(met_avg, elev_cat = ifelse(elev > 252, "high", "low"))
```

## #3 Use geom_violin to examine the wind speed and dew point by region
```{r}
met_avg <- na.omit(met_avg)
ggplot(met_avg, aes(x = 1, y = wind.sp)) +
  geom_violin(aes(fill = region), scale = "width", width = 0.8, na.rm = TRUE) +
  geom_violin(aes(x = 2, y = dew.point, fill = region), scale = "width", width = 0.8, na.rm = TRUE) +
  labs(x = NULL, y = "Value") +
  scale_x_continuous(breaks = c(1, 2), labels = c("Wind Speed", "Dew Point")) +
  facet_wrap(~region, nrow = 1) +  
  ggtitle("Wind Speed and Dew Point by Region")
```
It appears that wind speed and dew point have an inverse relationship.  Dew points are high when wind speed is lower.  It appears that in the southwest, there is a lot more dew overall compared to the other regions, having a large range.

## #4 Use geom_jitter with stat_smooth to examine the association between dew point and wind speed by region
```{r}
ggplot(met_avg, aes(x = dew.point, y = wind.sp, color = region, na.rm = TRUE)) +
  geom_jitter() +
  stat_smooth() +
  labs(x = "Dew Point", y = "Wind Speed") + 
  ggtitle("Association Between Dew Point and Wind Speed by Region")
```
In the graph, we can see that in the northwest, it has a lot of weather stations at the lower elevation levels.  At higher elevations, the northeast has a lot of weather stations but not as much as northwest at lower elevation.  In both elevation levels, southeast has the least amount of weather stations.

## #5 Use geom_bar to create barplots of the weather stations by elevation category colored by region
```{r}
ggplot(met_avg, aes(x = elev_cat, fill = region)) +
  geom_bar(position = "dodge", na.rm = TRUE) +
  scale_fill_brewer(palette = "PiYG") +
  labs(x = "Elevation Levels", y = "Count", fill = "Region") +
  ggtitle("Weather Stations by Elevation Levels and Region")
```
In the graph, we can see that in the northwest, it has a lot of weather stations at the lower elevation levels.  At higher elevations, the northeast has a lot of weather stations but not as much as northwest at lower elevation.  In both elevation levels, southeast has the least amount of weather stations.

## #6 Use stat_summary to examine mean dew point and wind speed by region with standard deviation error bars
```{r}
ggplot(met_avg, aes(x = region, y = dew.point)) +
  geom_point() + 
  geom_point(aes(y = wind.sp), color = "blue") + 
  stat_summary(
    fun.data = "mean_sdl",
    geom = "errorbar",
  ) +
  stat_summary(
    fun.data = "mean_sdl", 
    geom = "errorbar",
    aes(y = wind.sp),
    color = "blue"
  ) +
  labs(
    x = "Region",
    y = "Mean Value",
    title = "Mean Dew Point and Wind Speed by Region with Error Bars"
  )
```
Dew point is more spread across the values, with a large variance. Wind speed is more limited with a shorter range and has smaller variance in comparison.  The only ones where wind speeds are within the range of dew points are in the southeast and southwest.

## #7 Make a map showing the spatial trend in relative humidity in the US
```{r}
library(leaflet)
met_avg2 <- met[,.(rh, lat = lat, lon = lon), by=c("USAFID")]
met_avg2 <- met_avg2[!is.na(rh)]

rh.pal <- colorNumeric(c('blue','green', 'yellow', 'red'), domain = met_avg2$rh)
rh.pal

leaflet(met_avg2) %>%
 addProviderTiles('CartoDB.Positron') %>%
  addCircles(lat = ~lat, lng=~lon,
             label = ~paste0(round(rh,2), "%"), color = ~ rh.pal(rh),
             opacity = 1, fillOpacity = 1, radius = 5
             ) %>%
   addLegend(
    "bottomright",
    pal = rh.pal,
    values = ~rh,
    title = "Relative Humidity (%)",
    opacity = 0.7
  )

top_10 <- met_avg2 %>% filter(rank(-met_avg2$rh) <= 10)
leaflet(top_10) %>%
  addProviderTiles('CartoDB.Positron') %>%
  addMarkers(
    data = top_10, 
    lng = ~lon, 
    lat = ~lat,
    label = ~paste("Top 10: ", round(rh,2), "%"),
    options = markerOptions(icon = makeIcon(iconUrl = NULL))
  )
```
Across the US, the relative humidity appears to be warmer on the east coast(north and south) moreso than on the west coast(north and south).  Unfortunately, I was unable to get the markers for what is in the top 10, but I would assume it would be more towards the east since there is a lot of humidity there.

## #8 Use a ggplot extension
```{r}
library(ggforce)
ggplot(met_avg, aes(elev, elev_cat, colour = region)) +
  geom_point() +
  facet_zoom(x = region == "versicolor")
```
